---
title: "diannDataProcessing"
author: "Christopher Hughes"
format: html
editor: visual
---

## Details

This document describes processing of proteomics data from DIANN to compile a final protein set.

## Setting up the environment

These are packages you will need for this notebook. For exact versions used, please refer to the session info at the bottom of this notebook.

```{r}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
library('arrow')
library('iq')
library('DEqMS')
library('RColorBrewer')
library('ggrepel')
```

I want to set a base directory that we can use as a working space.

```{r}
##########################################################################################
baseRepository = 'E:/requests/1229'
```

## Data processing

Read in data from DIA-NN and process it to filtered sets. Do this for the serum and flowthrough data sets. 

```{r}
##########################################################################################
##we are using the data directly from DIA-NN because these were only single injections
diaData = read_tsv(paste(baseRepository,'/serumProfiling/diannDia1/report.pg_matrix.tsv',sep=''),
                   show_col_types = FALSE) %>%
  dplyr::filter(!grepl(';',Protein.Group)) %>%
  dplyr::filter(`N.Proteotypic.Sequences` > 0)

##parse the column names and update in the original tibble
colNameData = tibble('original' = colnames(diaData)[7:ncol(diaData)]) %>%
  dplyr::mutate('updated' = sub('.*Profiling_(.*)_dia1.*','\\1',original))
colnames(diaData)[7:ncol(diaData)] = colNameData$updated
colnames(diaData)[1:6] = c('accession',
                           'proteinName',
                           'gene',
                           'description',
                           'n_sequences',
                           'n_proteotypic')


##filter the data based on the number of observations of a protein
diaDataSet1 = diaData[,c(1:6,which(grepl('SP1',colnames(diaData))))]
diaDataSet1$numNA = rowSums(is.na(diaDataSet1[,7:51]))
write.table(diaDataSet1, paste(baseRepository, '/dataProcessing/serum_set1_parsedProteinData.tsv', sep = ''), 
            col.names = TRUE, row.names = FALSE, quote = FALSE, sep = '\t')
saveRDS(diaDataSet1, paste(baseRepository, '/dataProcessing/serum_set1_parsedProteinData.rds', sep = ''))

##plot the missing data
ggplot(diaDataSet1, aes(numNA, group=1)) +
  geom_histogram(binwidth = 1) +
  labs(title = 'number of missing values, dataset 1', x = 'number of missing values', y = 'count') +
  theme_classic()
ggsave(paste(baseRepository,'/dataProcessing/histogram_missingValues_dataset1.pdf',sep=''))

##filter the data
diaDataSet1Filtered = dplyr::filter(diaDataSet1, numNA <= 20)
write.table(diaDataSet1Filtered, paste(baseRepository, '/dataProcessing/serum_set1_parsedProteinDataFiltered.tsv', sep = ''), 
            col.names = TRUE, row.names = FALSE, quote = FALSE, sep = '\t')
saveRDS(diaDataSet1Filtered, paste(baseRepository, '/dataProcessing/serum_set1_parsedProteinDataFiltered.rds', sep = ''))


##repeat this process for the second set of data
##filter the data based on the number of observations of a protein
diaDataSet2 = diaData[,c(1:6,which(grepl('SP2',colnames(diaData))))]
diaDataSet2$numNA = rowSums(is.na(diaDataSet2[,7:51]))
write.table(diaDataSet2, paste(baseRepository, '/dataProcessing/serum_set2_parsedProteinData.tsv', sep = ''), 
            col.names = TRUE, row.names = FALSE, quote = FALSE, sep = '\t')
saveRDS(diaDataSet2, paste(baseRepository, '/dataProcessing/serum_set2_parsedProteinData.rds', sep = ''))

##plot the missing data
ggplot(diaDataSet2, aes(numNA, group=1)) +
  geom_histogram(binwidth = 1) +
  labs(title = 'number of missing values, dataset 2', x = 'number of missing values', y = 'count') +
  theme_classic()
ggsave(paste(baseRepository,'/dataProcessing/histogram_missingValues_dataset2.pdf',sep=''))

##filter the data
diaDataSet2Filtered = dplyr::filter(diaDataSet2, numNA <= 20)
write.table(diaDataSet2Filtered, paste(baseRepository, '/dataProcessing/serum_set2_parsedProteinDataFiltered.tsv', sep = ''), 
            col.names = TRUE, row.names = FALSE, quote = FALSE, sep = '\t')
saveRDS(diaDataSet2Filtered, paste(baseRepository, '/dataProcessing/serum_set2_parsedProteinDataFiltered.rds', sep = ''))




##########################################################################################
##below is for the flowthrough data
##we are using the data directly from DIA-NN because these were only single injections
diaData = read_tsv(paste(baseRepository,'/flowthroughProfiling/diannDia1/report.pg_matrix.tsv',sep=''),
                   show_col_types = FALSE) %>%
  dplyr::filter(!grepl(';',Protein.Group)) %>%
  dplyr::filter(`N.Proteotypic.Sequences` > 0)

##parse the column names and update in the original tibble
colNameData = tibble('original' = colnames(diaData)[7:ncol(diaData)]) %>%
  dplyr::mutate('updated' = sub('.*Profiling_(.*)_dia1.*','\\1',original))
colnames(diaData)[7:ncol(diaData)] = colNameData$updated
colnames(diaData)[1:6] = c('accession',
                           'proteinName',
                           'gene',
                           'description',
                           'n_sequences',
                           'n_proteotypic')


##filter the data based on the number of observations of a protein
diaDataSet1 = diaData[,c(1:6,which(grepl('SE',colnames(diaData))))]
diaDataSet1$numNA = rowSums(is.na(diaDataSet1[,7:51]))
write.table(diaDataSet1, paste(baseRepository, '/dataProcessing/flowthrough_set1_parsedProteinData.tsv', sep = ''), 
            col.names = TRUE, row.names = FALSE, quote = FALSE, sep = '\t')
saveRDS(diaDataSet1, paste(baseRepository, '/dataProcessing/flowthrough_set1_parsedProteinData.rds', sep = ''))

##plot the missing data
ggplot(diaDataSet1, aes(numNA, group=1)) +
  geom_histogram(binwidth = 1) +
  labs(title = 'number of missing values, dataset 1', x = 'number of missing values', y = 'count') +
  theme_classic()
ggsave(paste(baseRepository,'/dataProcessing/histogram_missingValues_dataset1.pdf',sep=''))

##filter the data
diaDataSet1Filtered = dplyr::filter(diaDataSet1, numNA <= 20)
write.table(diaDataSet1Filtered, paste(baseRepository, '/dataProcessing/flowthrough_set1_parsedProteinDataFiltered.tsv', sep = ''), 
            col.names = TRUE, row.names = FALSE, quote = FALSE, sep = '\t')
saveRDS(diaDataSet1Filtered, paste(baseRepository, '/dataProcessing/flowthrough_set1_parsedProteinDataFiltered.rds', sep = ''))
```

Take a look at the QC proteins we had spiked to examine instrument performance.

```{r}
##########################################################################################
##read in the data for the two sets
set1 = readRDS(paste(baseRepository, '/dataProcessing/serum_set1_parsedProteinData.rds', sep = '')) %>%
  dplyr::select(accession, SP1_1:SP1_45)
set2 = readRDS(paste(baseRepository, '/dataProcessing/serum_set2_parsedProteinData.rds', sep = '')) %>%
  dplyr::select(accession, SP2_1:SP2_45)

##combine and filter for our two QC proteins
qcSet = dplyr::left_join(set1, set2, by = c('accession')) %>%
  dplyr::filter(grepl('CONT117',accession) | grepl('CONT118',accession)) %>%
  dplyr::mutate('gene' = ifelse(grepl('CONT117',accession), 'rnase1','enolase')) %>%
  dplyr::select(-accession) %>%
  tidyr::pivot_longer(cols = SP1_1:SP2_45, names_to = 'dataset', values_to = 'signal')

##CONT117 is RNASE1, spiked into the digested samples
##CONT118 is enolase, spiked in pre-SP3
ggplot(qcSet, aes(dataset, log2(signal), color = gene)) +
  geom_point() +
  scale_y_continuous(limits = c(10,34), breaks = seq(0,40,4))



```

The code below is not used for this analysis, but I will leave it here in case I want to use it later.

Perform differential expression analysis on the data.

```{r}
##########################################################################################
##I am going to do one-by-one comparisons
processedData = readRDS(paste(baseRepository, '/dataProcessing/parsedProteinData.rds', sep = ''))

##read in the annotation details again
annotationDb = readRDS(paste(baseRepository, '/databases/2025-09-09-reviewed-UP000005640-spikein.fasta_index_trypticPeps.rds',sep='')) %>%
  dplyr::select(-species)

##rename the columns and setup the comparisons
#conditionsToCompare = c('latest_HO22')
#controlCondition = 'latest_EtOH'
#conditionsToCompare = c('t48h_HO22')
#controlCondition = 't48h_EtOH'
conditionsToCompare = c('t72h_HO22')
controlCondition = 't72h_EtOH'


#
for (i in 1:length(conditionsToCompare)){
  dat = processedData[,c(which(names(processedData) == 'accession'),
                         which(names(processedData) == 'gene'),      
                         which(names(processedData) == paste(controlCondition,'_A', sep = '')),
                         which(names(processedData) == paste(controlCondition,'_B', sep = '')),
                         which(names(processedData) == paste(controlCondition,'_C', sep = '')),
                         which(names(processedData) == paste(conditionsToCompare[i],'_A', sep = '')),
                         which(names(processedData) == paste(conditionsToCompare[i],'_B', sep = '')),
                         which(names(processedData) == paste(conditionsToCompare[i],'_C', sep = '')))]
  dat$samples123 = rowSums(is.na(dat[,3:5]))
  dat$samples456 = rowSums(is.na(dat[,6:8]))
  colnames(dat)[3:5] = c('cont_rep1','cont_rep2','cont_rep3')
  colnames(dat)[6:8] = c('treat_rep1','treat_rep2','treat_rep3')
  #
  datFiltered = as.data.frame(dat) %>%
    dplyr::filter(samples123 < 1 | samples456 < 1) %>%
    dplyr::mutate(cont_rep1 = ifelse(samples123 == 3, min(cont_rep1, na.rm = TRUE), cont_rep1),
                  cont_rep2 = ifelse(samples123 == 3, min(cont_rep2, na.rm = TRUE), cont_rep2),
                  cont_rep3 = ifelse(samples123 == 3, min(cont_rep3, na.rm = TRUE), cont_rep3)) %>%
    dplyr::mutate(treat_rep1 = ifelse(samples456 == 3, min(treat_rep1, na.rm = TRUE), treat_rep1),
                  treat_rep2 = ifelse(samples456 == 3, min(treat_rep2, na.rm = TRUE), treat_rep2),
                  treat_rep3 = ifelse(samples456 == 3, min(treat_rep3, na.rm = TRUE), treat_rep3))
  #
  rownames(datFiltered) = datFiltered$accession
  dat.log = datFiltered %>%
    dplyr::select(-accession, -gene, -samples123, -samples456)
  #dat.log = na.omit(datFiltered)
  #boxplot(dat.log, las = 2, main = 'total protein abundance')
  
  ##make a design table
  cond = as.factor(c(rep('cont',3),
                    rep('treat',3)))
  design = model.matrix(~0+cond)
  colnames(design) = gsub('cond', '', colnames(design))
  
  ##make the contrasts
  x = c('treat-cont')
  contrast = makeContrasts(contrasts = x, levels = design)
  fit1 = lmFit(dat.log, design)
  fit2 = contrasts.fit(fit1, contrasts = contrast)
  fit3 = eBayes(fit2)

  ##add information about PSM counts
  forPsm = processedData
  psm.count.table = data.frame(count = forPsm$n_peptides,
                               row.names = forPsm$accession)
  fit3$count = psm.count.table[rownames(fit3$coefficients),'count']
  fit4 = spectraCounteBayes(fit3)
  #VarianceBoxplot(fit4, n = 30, main = 'total proteome dataset', xlab = 'psm count')
  #VarianceScatterplot(fit4, main = 'total proteome dataset')
  
  ##extract the data
  #head(fit4$coefficients)
  psmAnnotate = dplyr::select(forPsm, accession, gene, n_peptides)
  valuesAnnotate = dplyr::select(datFiltered, accession, gene, samples123, samples456) %>%
    dplyr::rename(controlMissingValues = 'samples123',
                  conditionMissingValues = 'samples456')
  deqmsResults = tibble(outputResult(fit4, coef_col = 1)) %>%
    dplyr::rename(accession = 'gene') %>%
    dplyr::left_join(psmAnnotate) %>%
    dplyr::left_join(annotationDb) %>%
    dplyr::left_join(valuesAnnotate)
  #
  saveRDS(deqmsResults, paste(baseRepository, '/dataProcessing/dataset_deqms_', conditionsToCompare[i], '-', controlCondition, '.rds', sep = ''))
  write.table(deqmsResults, paste(baseRepository, '/dataProcessing/dataset_deqms_', conditionsToCompare[i], '-', controlCondition, '.tsv', sep = ''),
              col.names = TRUE, row.names = FALSE, sep = '\t', quote = FALSE)
  }
```
