---
title: "diannDataProcessing"
author: "Christopher Hughes"
format: html
editor: visual
---

## Details

This document describes processing of proteomics data from DIANN to compile a final protein set.

## Setting up the environment

These are packages you will need for this notebook. For exact versions used, please refer to the session info at the bottom of this notebook.

```{r}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
library('arrow')
library('iq')
library('DEqMS')
library('RColorBrewer')
library('ggrepel')
```

I want to set a base directory that we can use as a working space.

```{r}
##########################################################################################
baseRepository = 'E:/requests/1228/251117_1228_orbitrapLumos'
```

## Data processing

Read in data from DIA-NN and process it to filtered sets.

```{r}
##########################################################################################
##we are using the data directly from DIA-NN because these were only single injections
diaData = read_tsv(paste(baseRepository,'/diannDia1/report.pg_matrix.tsv',sep=''),
                   show_col_types = FALSE) %>%
  dplyr::filter(!grepl(';',Protein.Group)) %>%
  dplyr::filter(`N.Proteotypic.Sequences` > 0)

##parse the column names and update in the original tibble
colNameData = tibble('original' = colnames(diaData)[7:ncol(diaData)]) %>%
  dplyr::mutate('bmsId' = sub('.*turboId_(.*)_dia1.*','\\1',original),
                'bmsDate' = sub('.*orbitrapLumos\\\\(.*)_1228.*','\\1',original)) %>%
  dplyr::select(-original)
annotationSet = tibble('bmsId' = c(paste('sample',seq(1,18,1),sep=''), paste('sample',seq(1,12,1),sep='')),
                       'labId' = paste('sample',seq(1,30,1),sep=''),
                       'bmsDate' = c(rep('250808_1228',6),rep('250905_1228',12),rep('251117_1228',12))) %>%
  dplyr::left_join(colNameData)


colnames(diaData)[7:ncol(diaData)] = annotationSet$labId
colnames(diaData)[1:6] = c('accession',
                           'proteinName',
                           'gene',
                           'description',
                           'n_sequences',
                           'n_proteotypic')

#write the data
write_tsv(diaData, paste(baseRepository,'/diannDia1/report.pg_matrix.parsed.tsv',sep=''))

```


