---
title: "diannDataProcessing"
author: "Christopher Hughes"
format: html
editor: visual
---

## Details

This document describes processing of proteomics data from DIANN to compile a final protein set.

## Setting up the environment

These are packages you will need for this notebook. For exact versions used, please refer to the session info at the bottom of this notebook.

```{r}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
library('arrow')
library('iq')
library('DEqMS')
library('RColorBrewer')
library('ggrepel')
```

I want to set a base directory that we can use as a working space.

```{r}
##########################################################################################
baseRepository = 'E:/requests/1238/260218_1238_qexactive'
```

## Data processing

Read in data from DIA-NN and process with the IQ package.

```{r}
##########################################################################################
##first we can filter and parse the report data into a better format for IQ
##I had to do some extra stuff here because the sample ID was in a bad format
##if you are using DIA-NN version 1.9 or greater, you need to rename the Fragment.Quant column

#set the number of fractions you acquired
numberOfFractions = 1

#prefix of the raw data file
filePrefix = '260218_1238_lb_ipms_'

#process the data
raw = tibble()
for (i in seq(1,numberOfFractions,1)){
  dia1 = read_parquet(paste(baseRepository, '/diannDia',i,'/report.parquet',sep=''),) %>%
    dplyr::mutate(Run = sub(paste('.*',filePrefix,'(.*)_dia[123]',sep=''),'\\1', Run)) %>%
    #dplyr::filter(!grepl(';', Protein.Group)) %>%
    dplyr::mutate(fraction = i)
  #dia1$Intensities = paste(dia1$Fr.0.Quantity, dia1$Fr.1.Quantity, dia1$Fr.2.Quantity, 
  #                      dia1$Fr.3.Quantity, dia1$Fr.4.Quantity, dia1$Fr.5.Quantity, 
  #                      dia1$Fr.6.Quantity, dia1$Fr.7.Quantity, dia1$Fr.8.Quantity, 
  #                      dia1$Fr.9.Quantity, dia1$Fr.10.Quantity, dia1$Fr.11.Quantity, sep = ";")
  #
  raw = rbind(raw, dia1)
}

##save the data
write.table(raw, paste(baseRepository,'/dataProcessing/combinedFilteredReportAnnotated.tsv',sep=''), 
            col.names = TRUE, row.names = FALSE, quote = FALSE, sep = '\t')

##run the IQ analysis
iqResult = process_long_format(paste(baseRepository,'/dataProcessing/combinedFilteredReportAnnotated.tsv',sep=''),
                           output_filename = paste(baseRepository,'/dataProcessing/combinedFilteredReportAnnotatedIq.tsv',sep=''),
                           sample_id = "Run",
                           secondary_id = c("Precursor.Id", "fraction"),
                           primary_id = "Protein.Group",
                           intensity_col = "Precursor.Normalised",
                           intensity_col_sep = NULL,
                           annotation_col = c("Protein.Ids",
                                              "Protein.Names", 
                                              "Genes"),
                           filter_double_less = c("Q.Value" = "0.01",
                                                  "PG.Q.Value" = "0.05", 
                                                  "Lib.Q.Value" = "0.01", 
                                                  "Lib.PG.Q.Value" = "0.01"),
                           peptide_extractor = function(x) gsub("[0-9].*$", "", x))

##do the same for the human index
annotationDb = readRDS(paste(baseRepository, '/databases/2026-01-12-UP000000803-spikein.fasta_index_trypticPeps.rds',sep='')) %>%
  dplyr::select(-species)
  #dplyr::filter(!grepl('sp\\|V.*',gene))

##parse the IQ result
proData = read_tsv(paste(baseRepository,'/dataProcessing/combinedFilteredReportAnnotatedIq.tsv',sep=''), show_col_types = FALSE) %>%
  dplyr::rename(accession = 'Protein.Group') %>%
  dplyr::mutate(accession = sub('.*\\;','',accession)) %>%
  dplyr::select(-Genes,-peptides,-XIC) %>%
  dplyr::left_join(annotationDb) %>%
  dplyr::filter(!grepl('CONT',accession) & !grepl('\\;',Protein.Names))

##annotate the columns
#annoData = readxl::read_xlsx(paste(baseRepository,'/Sample_name_correlation-1172-202510.xlsx',sep=''),
#                             col_names = c('bmsId','userId'),
#                             skip = 1)

annoData = read_tsv(paste(baseRepository,'/bmsSampleMatrix.txt',sep=''),
                                  col_names = c('bmsId','userId'),
                                  skip = 1) %>%
  dplyr::mutate(bmsId = tolower(bmsId))
#

resultAnno = tibble(bmsId = colnames(proData)[6:(ncol(proData)-5)]) %>%
  dplyr::left_join(annoData)
#
colnames(proData)[6:(ncol(proData)-5)] = resultAnno$userId

##save the data
saveRDS(proData, paste(baseRepository, '/dataProcessing/parsedProteinData.rds', sep = ''))
write.table(proData, paste(baseRepository, '/dataProcessing/parsedProteinData.tsv', sep = ''), 
            col.names = TRUE, row.names = FALSE, quote = FALSE, sep = '\t')
```

Summarize the precursor file.

```{r}
##########################################################################################
##just want to summarize the precursor file for a user
prec = read_tsv(paste(baseRepository, '/diannDia1/report.pr_matrix.tsv',sep=''), show_col_types = FALSE) %>%
  dplyr::select("Protein.Group", Genes, "Precursor.Id",
                "E:\\requests\\1238\\260218_1238_qexactive\\260218_1238_lb_ipms_sample1_dia1.raw":"E:\\requests\\1238\\260218_1238_qexactive\\260218_1238_lb_ipms_sample8_dia1.raw") %>%
  dplyr::rename(accession = 'Protein.Group',
                gene = 'Genes',
                precId = 'Precursor.Id')


#rename the sample columns
colnames(prec)[4:11] = c(paste('sample',seq(1,8,1),sep=''))

#reshape the data to get counts
precReshape = prec %>%
  pivot_longer(cols = sample1:sample8, names_to = 'sample', values_to = 'intensity') %>%
  dplyr::mutate(pepCounts = ifelse(is.na(intensity), 0, 1)) %>%
  dplyr::select(-intensity, -precId) %>%
  dplyr::group_by(accession, gene, sample) %>%
  dplyr::summarise(nPeps = sum(pepCounts, na.rm = TRUE)) %>%
  tidyr::pivot_wider(id_cols = c('accession', 'gene'), names_from = 'sample', values_from = 'nPeps')

#save the data
write.table(precReshape, paste(baseRepository,'/dataProcessing/dataset_peptideCountsTable.tsv',sep=''),
            row.names = FALSE, quote = FALSE, sep = '\t')
```

Finished.

```{r}
sessionInfo()
```

