---
title: "diannDataProcessing"
author: "Christopher Hughes"
format: html
editor: visual
---

## Details

This document describes processing of proteomics data from DIANN to compile a final protein set.

## Setting up the environment

These are packages you will need for this notebook. For exact versions used, please refer to the session info at the bottom of this notebook.

```{r}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
library('readxl')
library('writexl')
```

I want to set a base directory that we can use as a working space.

```{r}
##########################################################################################
baseRepository = 'E:/requests/1164/comparisons'
```

## Data processing

Read in the data from Gilles and process the different comparisons of interest.

```{r}
##########################################################################################
##define the conditions
cellLines = c('Me',
              'Ra',
              'Re',
              'cII')
conditions = c('2_5','human')

##loop through the conditions
for (i in 1:length(cellLines)){
  ##define the columns of interest
  columnsOfInterest = paste(cellLines[i],seq(3,4,1),sep='')
  ##read in the data
  ##there is a lot of somewhat useless renaming of things that happens here
  rawData = readxl::read_xlsx(paste(baseRepository, '/combined_protein.xlsx',sep='')) %>%
    dplyr::select(Protein:`Combined Unique Spectral Count`,
                  paste('sample',columnsOfInterest,'human Intensity',sep=''),
                  paste('sample',columnsOfInterest,'2_5 Intensity',sep='')) %>%
    dplyr::rename(proteinLong = 'Protein',
                  accession = `Protein ID`,
                  uniprot = `Entry Name`,
                  gene = 'Gene',
                  proteinLength = `Protein Length`,
                  organism = 'Organism',
                  existence = `Protein Existence`,
                  description = 'Description',
                  proteinProbability = `Protein Probability`,
                  topPeptideProbability = `Top Peptide Probability`,
                  totalPeptides = `Combined Total Peptides`,
                  spectralCounts = `Combined Spectral Count`,
                  uniqueSpectralCounts = `Combined Unique Spectral Count`) %>%
    dplyr::select(-proteinLong,
                  -proteinLength,
                  -organism,
                  -existence,
                  -proteinProbability,
                  -topPeptideProbability,
                  -totalPeptides,
                  -spectralCounts,
                  -uniqueSpectralCounts) %>%
    dplyr::filter(!grepl('CONT', accession))
  
  ##count NA values
  rawData[rawData==0] = NA
  rawData$control_NAs = rowSums(is.na(rawData[,5:6]))
  rawData$condition_NAs = rowSums(is.na(rawData[,7:8]))
  
  ##filter rows with only NAs
  rawDataFiltered1 = dplyr::filter(rawData, control_NAs == 0 | condition_NAs == 0)
  
  ##calculate overall values
  processedData = rawDataFiltered1 %>%
    dplyr::mutate(avg_control = rowMeans(rawDataFiltered1[,5:6],na.rm=TRUE),
                  avg_condition = rowMeans(rawDataFiltered1[,7:8],na.rm=TRUE)) %>%
    dplyr::mutate(condition_control_fc = avg_condition/avg_control) %>%
    dplyr::mutate(log2Fc = log2(condition_control_fc))
  
  ##assign a fake value to the missing comparison values
  processedDataWithMissing = processedData %>%
    dplyr::mutate(log2Fc = ifelse(condition_NAs == 0 & control_NAs == 2, 10, log2Fc))
  
  ##save the data
  writexl::write_xlsx(processedData, paste(baseRepository,'/dataset_',cellLines[i],'_condition-versus-control.xlsx',sep=''))
}



```
