---
title: "fairn lab data processing"
author: "Christopher Hughes"
format: html
---

## Details

This document describes preparing CRISPR count data to be input into the Broad CRISPR analysis tool.

## Setting up the environment

These are packages you will need for this notebook. For exact versions used, please refer to the session info at the bottom of this notebook.

```{r}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
```

I want to set a base directory that we can use for our analysis.

```{r}
##########################################################################################
baseRepository = 'E:/requests/1182/250814_1182_crisprScreen'
```

## Data processing

Process the fragpipe data.

```{r}
##########################################################################################
##barcodes
#barcodes = c('AAGAGGCA-TACTCCTT',
#             'AGGCAGAA-AGAGGATA',
#             'ATCTCAGG-AGGCTTAG', 
#             'CGAGGCTG-TATGCAGT', 
#             'CGTACTAG-ATAGAGAG', 
#             'CGTACTAG-CGGAGAGA', 
#             'CTCTCTAC-TATGCAGT', 
#             'GCTCATGA-AGGCTTAG', 
#             'GGACTCCT-CTCCTTAC', 
#             'GTAGAGGA-TACTCCTT', 
#             'TAAGGCGA-ATAGAGAG', 
#             'TAAGGCGA-ATTAGACG', 
#             'TAGGCATG-CTCCTTAC', 
#             'TCCTGAGC-AGAGGATA')
barcodes = read_tsv(paste(baseRepository,'/haifengReference.txt',sep=''), show_col_types = FALSE)

##reference file
brunelloRef = read_csv(paste(baseRepository,'/brunelloGuidesModified.csv',sep=''), show_col_types = FALSE,
                       col_names = c('guide','seq','gene'),
                       skip = 1)

#loop through the barcodes and transform the guide-counter result into scores
combinedCounts = tibble()
for (i in 1:length(barcodes$barcode)){
  countTemp = read_tsv(paste(baseRepository,'/guideCounterOutput/',barcodes$barcode[i],'.counts.txt',sep=''),
                       col_names = c('guide','gene','counts'),
                       skip = 1,
                       show_col_types = FALSE) %>%
    dplyr::mutate(score = log2(((counts / sum(counts, na.rm = TRUE))*1e6)+1)) %>%
    dplyr::left_join(brunelloRef) %>%
    dplyr::mutate(condition = barcodes$sample[i]) %>%
    dplyr::select(condition,seq,score)
  
  #combine
  combinedCounts = rbind(combinedCounts, countTemp)
}

#make into wide format
countsWide = combinedCounts %>%
  tidyr::pivot_wider(id_cols = 'seq', names_from = 'condition', values_from = 'score')

#output files
write_tsv(countsWide, paste(baseRepository,'/guideCounterOutput/totalCountMatrix_broadInput.txt',sep=''))

#total count matrix
totalCounts = countsWide %>%
  dplyr::left_join(brunelloRef) %>%
  dplyr::relocate(guide, seq, gene, 
                  ABR_T0_1, ABR_T0_2, ABR_T0_3,
                  ABR_B0_1, ABR_B0_2, ABR_B0_3,
                  ABR_B40_1, ABR_B40_2, ABR_B40_3,
                  ABR_B_Se_1, ABR_B_Se_2, ABR_B_Se_3,
                  A_IL_L20, A_IL_H20)
write_tsv(totalCounts, paste(baseRepository,'/guideCounterOutput/totalCountMatrix.txt',sep=''))

#comparison for +/-BSO
comparisonData = countsWide %>%
  dplyr::select(seq,
                  ABR_B0_1,ABR_B0_2,ABR_B0_3,
                  ABR_B40_1,ABR_B40_2,ABR_B40_3) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(cond1_mean = mean(c_across(ABR_B0_1:ABR_B0_3), na.rm = TRUE),
                cond2_mean = mean(c_across(ABR_B40_1:ABR_B40_3), na.rm = TRUE)) %>%
  dplyr::mutate(fcScore = cond2_mean - cond1_mean) %>%
  dplyr::select(seq, fcScore)
write_tsv(comparisonData, paste(baseRepository,'/guideCounterOutput/comparisonMatrix_B40-vs-B0_broadInput.txt',sep=''))
  
  
#comparison for +/-Se
comparisonData = countsWide %>%
  dplyr::select(seq,
                  ABR_B_Se_1,ABR_B_Se_2,ABR_B_Se_3,
                  ABR_B40_1,ABR_B40_2,ABR_B40_3) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(cond1_mean = mean(c_across(ABR_B_Se_1:ABR_B_Se_3), na.rm = TRUE),
                cond2_mean = mean(c_across(ABR_B40_1:ABR_B40_3), na.rm = TRUE)) %>%
  dplyr::mutate(fcScore = cond2_mean - cond1_mean) %>%
  dplyr::select(seq, fcScore)
write_tsv(comparisonData, paste(baseRepository,'/guideCounterOutput/comparisonMatrix_B40-vs-Se_broadInput.txt',sep=''))  
  
  
  
  
  
  
  
                
```