---
title: "IP sample comparisons"
author: "Christopher Hughes"
format: html
editor: visual
---

## Details

This document describes processing of FragPipe results to compare IP samples.

## Setting up the environment

These are packages you will need for this notebook. For exact versions used, please refer to the session info at the bottom of this notebook.

```{r}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
```

I want to set a base directory that we can use for our analysis.

```{r}
##########################################################################################
baseRepository = 'E:/requests/1192/240522_orbitrapLumos'
```

## Data processing

Process the FragPipe data for all of the samples.

```{r}
##########################################################################################
##read in the results files
sampleNames = c('EVb','EVc','EVd','K3b','K3c','K3d','K3VGb','K3VGc','K3VGd','VGb','VGc','VGd')
proCombined = tibble()
for (i in 1:length(sampleNames)){
  tempPro = read_tsv(paste(baseRepository, '/fragpipeFirstPass/',sampleNames[i],'/proteinProcessed.tsv',sep=''),show_col_types = FALSE) %>%
    dplyr::filter(!grepl('CONT',gene) & !grepl('KRT',gene) & uniquePeptides > 1) %>%
    dplyr::mutate(dataset = sampleNames[i]) %>%
    dplyr::select(dataset, accession, gene, pringleId, uniquePeptides, razorPeptides, detectablePeptides, iBAQ)
  #
  proCombined = rbind(proCombined, tempPro)
}


##reshape the data
proWide = proCombined %>%
  dplyr::select(dataset:pringleId, iBAQ) %>%
  tidyr::pivot_wider(id_cols = c('accession','gene','pringleId'), names_from = 'dataset', values_from = 'iBAQ')
proWide$naSet1 = rowSums(is.na(proWide[,4:6]))
proWide$naSet2 = rowSums(is.na(proWide[,7:9]))
proWide$naSet3 = rowSums(is.na(proWide[,10:12]))
proWide$naSet4 = rowSums(is.na(proWide[,13:15]))
##filter
proFilter = proWide %>%
  dplyr::filter(naSet1 < 2 | naSet2 < 2 | naSet3 < 2 | naSet4 < 2) %>%
  dplyr::rename(virusAccession = 'pringleId') %>%
  dplyr::select(accession:VGd)

##save the data
write_tsv(proFilter, paste(baseRepository, '/dataProcessing/dataset_combinedFilteredProteinData.tsv',sep=''))

write_tsv()
```


