---
title: "diannDataProcessing"
author: "Christopher Hughes"
format: html
editor: visual
---

## Details

This document describes processing of proteomics data from DIANN to compile a final protein set.

## Setting up the environment

These are packages you will need for this notebook. For exact versions used, please refer to the session info at the bottom of this notebook.

```{r}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
library('arrow')
library('iq')
library('DEqMS')
library('RColorBrewer')
library('ggrepel')
```

I want to set a base directory that we can use as a working space.

```{r}
##########################################################################################
baseRepository = 'E:/requests/1192/250904_1192_orbitrapLumos'
```

## Data processing

Read in data from DIA-NN and process with the IQ package.

```{r}
##########################################################################################
##define the sample names
sampleDetails = readxl::read_xlsx(paste(baseRepository,'/Sample_name_correlation-1192-20250825.xlsx',sep=''),
                                  col_names = c('bmsId','userId'),
                                  skip = 1)

##read in the annotation index
annotationDb = readRDS(paste(baseRepository, '/databases/2025-09-09-reviewed-UP000005640-spikein.fasta_index_trypticPeps.rds',sep='')) %>%
  dplyr::select(-species) %>%
  dplyr::filter(!grepl('sp\\|V.*',gene))



##read in the results file from DIA-NN
dia1 = read_tsv(paste(baseRepository,'/diannDia1/report.pg_matrix.tsv',sep=''),show_col_types = FALSE,
                col_names = c('accession','proteinNames','gene','description','n_sequences','n_proteotypicSequences',
                              sampleDetails$userId),
                skip = 1) %>%
  dplyr::filter(n_proteotypicSequences > 0 & !grepl(';',accession)) %>%
  dplyr::left_join(annotationDb) %>%
  dplyr::mutate(description = ifelse(grepl('VIRUS',accession), proteinNames, description)) %>%
  dplyr::select(-proteinNames)
  
##save the data
saveRDS(dia1, paste(baseRepository, '/dataProcessing/parsedProteinData.rds', sep = ''))
write.table(dia1, paste(baseRepository, '/dataProcessing/parsedProteinData.tsv', sep = ''), 
            col.names = TRUE, row.names = FALSE, quote = FALSE, sep = '\t')
```

Perform differential expression analysis on the data.

```{r}
##########################################################################################
##I am going to do one-by-one comparisons
processedData = readRDS(paste(baseRepository, '/dataProcessing/parsedProteinData.rds', sep = ''))

##read in the annotation details again
annotationDb = readRDS(paste(baseRepository, '/databases/2025-09-09-reviewed-UP000005640-spikein.fasta_index_trypticPeps.rds',sep='')) %>%
  dplyr::select(-species) %>%
  dplyr::filter(!grepl('sp\\|V.*',gene))

##rename the columns and setup the comparisons
#conditionsToCompare = c('WT_24h','WT_48h','WT_72h')
#controlCondition = 'WT_LAT'
conditionsToCompare = c('deltaK3_24h','deltaK3_48h','deltaK3_72h')
controlCondition = 'deltaK3_LAT'

#
for (i in 1:length(conditionsToCompare)){
  dat = processedData[,c(which(names(processedData) == 'accession'),
                         which(names(processedData) == 'gene'),      
                         which(names(processedData) == paste('A_',controlCondition, sep = '')),
                         which(names(processedData) == paste('B_',controlCondition, sep = '')),
                         which(names(processedData) == paste('C_',controlCondition, sep = '')),
                         which(names(processedData) == paste('A_',conditionsToCompare[i], sep = '')),
                         which(names(processedData) == paste('B_',conditionsToCompare[i], sep = '')),
                         which(names(processedData) == paste('C_',conditionsToCompare[i], sep = '')))]
  dat$samples123 = rowSums(is.na(dat[,3:5]))
  dat$samples456 = rowSums(is.na(dat[,6:8]))
  colnames(dat)[3:5] = c('cont_rep1','cont_rep2','cont_rep3')
  colnames(dat)[6:8] = c('treat_rep1','treat_rep2','treat_rep3')
  #
  datFiltered = as.data.frame(dat) %>%
    dplyr::filter(samples123 < 1 | samples456 < 1) %>%
    dplyr::mutate(cont_rep1 = ifelse(samples123 == 3, min(cont_rep1, na.rm = TRUE), cont_rep1),
                  cont_rep2 = ifelse(samples123 == 3, min(cont_rep2, na.rm = TRUE), cont_rep2),
                  cont_rep3 = ifelse(samples123 == 3, min(cont_rep3, na.rm = TRUE), cont_rep3)) %>%
    dplyr::mutate(treat_rep1 = ifelse(samples456 == 3, min(treat_rep1, na.rm = TRUE), treat_rep1),
                  treat_rep2 = ifelse(samples456 == 3, min(treat_rep2, na.rm = TRUE), treat_rep2),
                  treat_rep3 = ifelse(samples456 == 3, min(treat_rep3, na.rm = TRUE), treat_rep3))
  #
  rownames(datFiltered) = datFiltered$accession
  dat.log = datFiltered %>%
    dplyr::select(-accession, -gene, -samples123, -samples456)
  #dat.log = na.omit(datFiltered)
  #boxplot(dat.log, las = 2, main = 'total protein abundance')
  
  ##make a design table
  cond = as.factor(c(rep('cont',3),
                    rep('treat',3)))
  design = model.matrix(~0+cond)
  colnames(design) = gsub('cond', '', colnames(design))
  
  ##make the contrasts
  x = c('treat-cont')
  contrast = makeContrasts(contrasts = x, levels = design)
  fit1 = lmFit(dat.log, design)
  fit2 = contrasts.fit(fit1, contrasts = contrast)
  fit3 = eBayes(fit2)

  ##add information about PSM counts
  forPsm = processedData
  psm.count.table = data.frame(count = forPsm$n_sequences,
                               row.names = forPsm$accession)
  fit3$count = psm.count.table[rownames(fit3$coefficients),'count']
  fit4 = spectraCounteBayes(fit3)
  #VarianceBoxplot(fit4, n = 30, main = 'total proteome dataset', xlab = 'psm count')
  #VarianceScatterplot(fit4, main = 'total proteome dataset')
  
  ##extract the data
  #head(fit4$coefficients)
  psmAnnotate = dplyr::select(forPsm, accession, gene, n_sequences, n_proteotypicSequences)
  valuesAnnotate = dplyr::select(datFiltered, accession, gene, samples123, samples456) %>%
    dplyr::rename(controlMissingValues = 'samples123',
                  conditionMissingValues = 'samples456')
  deqmsResults = tibble(outputResult(fit4, coef_col = 1)) %>%
    dplyr::rename(accession = 'gene') %>%
    dplyr::left_join(psmAnnotate) %>%
    dplyr::left_join(annotationDb) %>%
    dplyr::left_join(valuesAnnotate)
  #
  saveRDS(deqmsResults, paste(baseRepository, '/dataProcessing/dataset_deqms_', conditionsToCompare[i], '-', controlCondition, '.rds', sep = ''))
  write.table(deqmsResults, paste(baseRepository, '/dataProcessing/dataset_deqms_', conditionsToCompare[i], '-', controlCondition, '.tsv', sep = ''),
              col.names = TRUE, row.names = FALSE, sep = '\t', quote = FALSE)
  }
```
