---
title: "Parsing MQ Data"
author: "Christopher Hughes"
format: html
editor: visual
---

## Details

This document describes processing AP-MS results to get enrichment values. There was some confusion about replicate names, some details from the client here:

-   1 - EV or EV pIC - empty vector or empty vector with poly IC - no problem with those
-   2 - *-delta (+/- pIC) - this was how Rory labeled the truncated N samples (N*M210) - we believe that for n=2 and 3 you thought this was full-length N (NFL)
-   3 - FL-delta delta - this was how Rory labeled the NFL samples - we believe that you thought this was N\*M210 for n=2 and n=3
-   In sum: NFL (aa 1-209 intact) = delta delta N\*M210 (lacking aa 1-209) = \*delta"

## Setting up the environment

These are packages you will need for this notebook. For exact versions used, please refer to the session info at the bottom of this notebook.

```{r}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
```

I want to set a base directory that we can use for our analysis.

```{r}
##########################################################################################
baseRepository = 'E:/requests/1219/250704_1219_orbitrapLumos'
```

## Data processing

First we want to check the peptide data to make sure the truncation of the protein is being detected as expected. There may be a replicate where peptides belonging to what should be a deleted region are detected.

Process the peptides file from MaxQuant.

```{r}
##########################################################################################
##read the peptides file and subset it to only contain peptides before AA position 210 for the bait protein
pepData = read_tsv(paste(baseRepository,'/combined/txt/peptides.txt',sep=''),
                  show_col_types = FALSE) %>%
  dplyr::filter(grepl('P0DTC9', `Leading razor protein`)) %>%
  dplyr::select(Sequence,
                `Leading razor protein`,
                `Start position`,
                `End position`,
                Score:Reverse) %>%
  dplyr::filter(`End position` < 211) %>%
  dplyr::select(-Score, -Reverse, -Intensity)

##read in your sample annotation data and parse into appropriate column names
annoData = read_tsv(paste(baseRepository,'/bmsSampleMatrix.txt',sep=''),show_col_types = FALSE)
mqColumns = tibble('mqNames' = colnames(pepData)[5:(nrow(annoData)+4)]) %>%
  dplyr::mutate(bmsId = paste('sample',sub('.*_sample(.*)','\\1',mqNames),sep='')) %>%
  dplyr::left_join(annoData)
colnames(pepData)[5:(nrow(annoData)+4)] = paste('psm_',mqColumns$sampleName, sep='')

##repeat this for the intensity columns
annoData = read_tsv(paste(baseRepository,'/bmsSampleMatrix.txt',sep=''),show_col_types = FALSE)
mqColumns = tibble('mqNames' = colnames(pepData)[23:(nrow(annoData)+22)]) %>%
  dplyr::mutate(bmsId = paste('sample',sub('.*_sample(.*)','\\1',mqNames),sep='')) %>%
  dplyr::left_join(annoData)
colnames(pepData)[23:(nrow(annoData)+22)] = paste('intensity_',mqColumns$sampleName, sep='')

##only the NFL samples should have signal in that first region, so we can get rid of these
pepDataFiltered = pepData %>%
  dplyr::select(!contains('N-FL')) 

#save these data
write.table(pepDataFiltered, paste(baseRepository,'/dataProcessing/dataset_peptideData_maxquant.tsv',sep=''),
            quote = FALSE, sep = '\t', row.names = FALSE)
```

Based on the above file, it looks like the samples 'EV_PIC_1' and 'N_M210_PIC_1' both have signal where they shouldn't. So, we will exclude this from the protein summarization.

Process the protein groups file.

```{r}
##########################################################################################
##read the proteinGroups file
pgData = read_tsv(paste(baseRepository,'/combined/txt/proteinGroups.txt',sep=''),
                  show_col_types = FALSE) %>%
  dplyr::filter(!grepl('CONT',`Protein IDs`) & !grepl('REV',`Protein IDs`)) %>%
  dplyr::select(`Protein IDs`,
                `Protein names`,
                `Gene names`,
                `Number of proteins`,
                Peptides:`Unique peptides`,
                iBAQ:`MS/MS count`,
                `Sequence length`) %>%
  dplyr::select(-iBAQ, -`MS/MS count`) %>%
  dplyr::rename('originalAccession' = `Protein IDs`,
                'description' = `Protein names`,
                'originalSymbol' = `Gene names`,
                'numberOfProteins' = `Number of proteins`,
                'totalPeptides' = Peptides,
                'razorPeptides' = `Razor + unique peptides`,
                'uniquePeptides' = `Unique peptides`,
                'length' = `Sequence length`) %>%
  dplyr::mutate(accession = sub(';.*','',originalAccession),
                symbol = sub(';.*','',originalSymbol))


##read in your sample annotation data and parse into appropriate column names
annoData = read_tsv(paste(baseRepository,'/bmsSampleMatrix.txt',sep=''),show_col_types = FALSE)
mqColumns = tibble('mqNames' = colnames(pgData)[8:(nrow(annoData)+7)]) %>%
  dplyr::mutate(bmsId = paste('sample',sub('.*_sample(.*)','\\1',mqNames),sep='')) %>%
  dplyr::left_join(annoData)
colnames(pgData)[8:(nrow(annoData)+7)] = paste('ibaq_',mqColumns$sampleName,sep='')

##repeat the above process for the LFQ columns
annoData = read_tsv(paste(baseRepository,'/bmsSampleMatrix.txt',sep=''),show_col_types = FALSE)
mqColumns = tibble('mqNames' = colnames(pgData)[26:(nrow(annoData)+25)]) %>%
  dplyr::mutate(bmsId = paste('sample',sub('.*_sample(.*)','\\1',mqNames),sep='')) %>%
  dplyr::left_join(annoData)
colnames(pgData)[26:(nrow(annoData)+25)] = paste('lfq_',mqColumns$sampleName,sep='')


##repeat the above process for the ms/ms count columns
annoData = read_tsv(paste(baseRepository,'/bmsSampleMatrix.txt',sep=''),show_col_types = FALSE)
mqColumns = tibble('mqNames' = colnames(pgData)[44:(nrow(annoData)+43)]) %>%
  dplyr::mutate(bmsId = paste('sample',sub('.*_sample(.*)','\\1',mqNames),sep='')) %>%
  dplyr::left_join(annoData)
colnames(pgData)[44:(nrow(annoData)+43)] = paste('msms_',mqColumns$sampleName,sep='')

##remove the two samples that are giving unreliable data
pgDataFiltered = pgData %>%
  dplyr::select(!contains('EV_PIC_1')) %>%
  dplyr::select(!contains('N_M210_PIC_1')) %>%
  dplyr::arrange(-totalPeptides) %>%
  dplyr::mutate(description = ifelse(grepl('P0DTC9', accession), 'N protein', description),
                symbol = ifelse(grepl('P0DTC9', accession), 'N', symbol))

##write the data
write_tsv(pgDataFiltered, paste(baseRepository,'/dataProcessing/dataset_proteinData_maxquant.tsv',sep=''))
```

Prepare the data for saint-express analysis.

```{r}
##########################################################################################
##read in the protein data
pgData = read_tsv(paste(baseRepository,'/dataProcessing/dataset_proteinData_maxquant.tsv',sep=''),
                  show_col_types = FALSE)


##create the 'bait' file
bait = tibble('ipName' = c(paste('EV_',seq(1,3,1),sep=''),
                           paste('N-FL_',seq(1,3,1),sep='')),
              'baitName' = c(paste('EV_',seq(1,3,1),sep=''),
                             rep('N',3)),
              'testType' = c(rep('C',3),
                             rep('T',3)))


##create the 'prey' file
#this needs to be unique, so first check for non-unique entries
pgData[which(duplicated(pgData$symbol)),'symbol']
pgData[which(duplicated(pgData$accession)),'accession']
#
prey = tibble('preyName' = pgData$accession,
              'preyLength' = pgData$length,
              'preySymbol' = pgData$symbol)


##perform normalization on the IPs of interest
rawData = pgData %>%
  dplyr::select(contains(paste('lfq_',bait$ipName[4],sep='')),
                contains(paste('lfq_',bait$ipName[5],sep='')),
                contains(paste('lfq_',bait$ipName[6],sep='')))
intensityTotals = colSums(rawData) #get the total intensity for each sample
normFactors = intensityTotals / min(intensityTotals) #determine the normalization factor
#design a new bait matrix with the normalization factors
baitNorm = bait %>%
  dplyr::mutate(normFactors = c(rep(1,3),
                                normFactors))


##create the 'interaction' file
interaction = tibble()
for (i in 1:length(bait$ipName)){
  ##
  tmpData = pgData %>%
    dplyr::select(accession,
                  contains(paste('lfq_',bait$ipName[i],sep=''))) %>%
    dplyr::rename(lfqAbundance = paste('lfq_',bait$ipName[i],sep='')) %>%
    dplyr::filter(lfqAbundance > 0) %>%
    dplyr::mutate(lfqAbundance = lfqAbundance / baitNorm$normFactors[i]) %>%
    dplyr::mutate(lfqAbundance = log2(lfqAbundance)) %>%
    dplyr::mutate(ipName = bait$ipName[i],
                  baitName = bait$baitName[i]) %>%
    dplyr::select(ipName,
                  baitName,
                  accession,
                  lfqAbundance)
  #
  interaction = rbind(interaction, tmpData)
}

##save these files
write.table(bait, paste(baseRepository,'/dataProcessing/bait.dat',sep=''),
            row.names = FALSE, col.names = FALSE, quote = FALSE, sep = '\t')
#
write.table(prey, paste(baseRepository,'/dataProcessing/prey.dat',sep=''),
            row.names = FALSE, col.names = FALSE, quote = FALSE, sep = '\t')
#
write.table(interaction, paste(baseRepository,'/dataProcessing/interaction.dat',sep=''),
            row.names = FALSE, col.names = FALSE, quote = FALSE, sep = '\t')


##now we need to run saint-express, I did this using the command below
#/mnt/c/softwareTools/proteomics/saint-express/SAINTexpress_v3.6.3__2018-03-09/SAINTexpress_v3.6.3__2018-03-09/Precompiled_binaries/Linux64$ ./SAINTexpress-int interaction.dat prey.dat bait.dat

##looking at the saint data from above highlighted we likely need some normalization in the IP samples, so I went back to apply this in the interaction file
#we don't perform normalization on the EV samples
```















