---
title: "peptideDataProcessing"
author: "Christopher Hughes"
format: html
editor: visual
---

## Details

This document describes processing of data from Fairn lab in order to look at GLUD1.

## Setting up the environment

These are packages you will need for this notebook. For exact versions used, please refer to the session info at the bottom of this notebook.

```{r}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
```

I want to set a base directory that we can use for our analysis.

```{r}
##########################################################################################
baseRepository = 'D:/requests/1178/240209_orbitrapLumos'
```

## Data processing

Read in the psm data and process into a set for GLUD1.In the first part of this I do normalization of the intensity data.

```{r}
##########################################################################################
##read the PSM data
sample1 = read_tsv(paste(baseRepository, '/fragpipeAnalysis/control1/psm.tsv',sep=''), show_col_types = FALSE) %>%
  dplyr::select(Intensity)
sample2 = read_tsv(paste(baseRepository, '/fragpipeAnalysis/control2/psm.tsv',sep=''), show_col_types = FALSE) %>%
  dplyr::select(Intensity)
sample3 = read_tsv(paste(baseRepository, '/fragpipeAnalysis/palm1/psm.tsv',sep=''), show_col_types = FALSE) %>%
  dplyr::select(Intensity)
sample4 = read_tsv(paste(baseRepository, '/fragpipeAnalysis/palm2/psm.tsv',sep=''), show_col_types = FALSE) %>%
  dplyr::select(Intensity)
target = mean(c(colSums(sample1, na.rm = TRUE), colSums(sample2, na.rm = TRUE), colSums(sample3, na.rm = TRUE), colSums(sample4, na.rm = TRUE)))
normFacs = target/colSums(sample1, na.rm = TRUE)
sample1Norm = sweep(sample1, 2, normFacs, FUN = "*")
normFacs = target/colSums(sample2, na.rm = TRUE)
sample2Norm = sweep(sample2, 2, normFacs, FUN = "*")
normFacs = target/colSums(sample3, na.rm = TRUE)
sample3Norm = sweep(sample3, 2, normFacs, FUN = "*")
normFacs = target/colSums(sample4, na.rm = TRUE)
sample4Norm = sweep(sample4, 2, normFacs, FUN = "*")
sampleNormData = list(sample1Norm, sample2Norm, sample3Norm, sample4Norm)

##read PSM data
samplesToProcess = c('control1','control2','palm1','palm2')
peptideData = tibble()

##
for (i in 1:length(samplesToProcess)){
  psmTemp = read_tsv(paste(baseRepository, '/fragpipeAnalysis/',samplesToProcess[i],'/psm.tsv',sep=''), show_col_types = FALSE) %>%
    dplyr::mutate(intensityNorm = sampleNormData[[i]]$Intensity) %>%
    dplyr::select(Peptide, `Protein Start`, `Protein End`, Intensity, `Assigned Modifications`, `Protein ID`, Gene, intensityNorm) %>%
    dplyr::filter(Gene == 'GLUD1' & grepl('C',Peptide) & Intensity > 0) %>%
    dplyr::rename(accession = `Protein ID`,
                  gene = "Gene",
                  sequence = "Peptide",
                  proStart = "Protein Start",
                  proEnd = "Protein End",
                  intensity = "Intensity",
                  normIntensity = "intensityNorm",
                  mods = "Assigned Modifications") %>%
    dplyr::group_by(accession, gene, sequence, proStart, proEnd, mods) %>%
    dplyr::mutate(psms = 1) %>%
    dplyr::summarise(medIntensity = median(intensity, na.rm = TRUE),
                     medNormIntensity = median(normIntensity, na.rm = TRUE),
                     numPsms = sum(psms, na.rm = TRUE)) %>%
    dplyr::mutate(sampleId = samplesToProcess[i])
  #
  peptideData = rbind(peptideData, psmTemp)
}

##save these data
saveRDS(peptideData, paste(baseRepository, '/cysteineProfiling/dataset_parsedPeptideData.rds',sep=''))
write.table(peptideData, paste(baseRepository, '/cysteineProfiling/dataset_parsedPeptideData.tsv',sep=''),
            quote = FALSE, col.names = TRUE, row.names = FALSE, sep = '\t')

##make wide
peptideWide = peptideData %>%
  dplyr::ungroup() %>%
  dplyr::select(sequence, sampleId, numPsms, medIntensity, medNormIntensity) %>%
  tidyr::pivot_wider(id_cols = 'sampleId', names_from = 'sequence', values_from = c('numPsms','medIntensity','medNormIntensity'))
```

## Session info

```{r}
##########################################################################################
sessionInfo()
```

