---
title: "databaseProcessing"
author: "Christopher Hughes"
format: html
editor: visual
---

## Details

This document describes processing a database file to so that we can combine it with a human database for processing with DIANN.

## Setting up the environment

These are packages you will need for this notebook. For exact versions used, please refer to the session info at the bottom of this notebook.

```{r}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
library('Biostrings')
```

I want to set a base directory that we can use for our analysis.

```{r}
##########################################################################################
baseRepository = 'D:/requests/231213_1172/project_3d2'
```

## Data processing

Process the virus fasta database.

```{r}
##########################################################################################
fastaDb = readAAStringSet(paste(baseRepository,'/ncbiProtein_GQ994935.1_jan2024.fasta',sep=''))

##define an index for each entry
fastaIndex = tibble('metadata' = names(fastaDb)) %>%
  mutate(accession = paste('sp|', sub('(.*)\\.[1] [KO].*','\\1', metadata), '|VIRUS', seq(1,nrow(.),1), '_HUMAN ', 'GN=VIRUS', seq(1,nrow(.),1), sep=''))

##save the fasta index for use later on
saveRDS(fastaIndex, paste(baseRepository,'/ncbiViralDatabaseIndex.rds',sep=''))

##edit the fastadb names
fastaDbEdited = fastaDb
names(fastaDbEdited) = fastaIndex$accession

##write the new fasta database
writeXStringSet(fastaDbEdited, paste(baseRepository,'/ncbiViralDatabaseReannotated.fasta',sep=''))
```

Process the crap database.

```{r}
##########################################################################################
fastaDb = readAAStringSet(paste(baseRepository,'/crap_2024_01_15.fasta',sep=''))

##define an index for each entry
fastaIndex = tibble('metadata' = names(fastaDb)) %>%
  dplyr::mutate(accession = paste('sp|CONT', seq(1,nrow(.),1), '|', sub('sp\\|(.*)\\|','\\1', metadata), ' GN=', sub('sp\\|(.*)_.*','\\1', metadata), sep=''))

##save the fasta index for use later on
saveRDS(fastaIndex, paste(baseRepository,'/crap_2024_01_15_Index.rds',sep=''))

##edit the fastadb names
fastaDbEdited = fastaDb
names(fastaDbEdited) = fastaIndex$accession
toKeep = data.frame(name = names(fastaDbEdited)[which(!grepl('HUMAN', names(fastaDbEdited)))])

##write the new fasta database
writeXStringSet(fastaDbEdited[toKeep$name], paste(baseRepository,'/crapDatabaseReannotated.fasta',sep=''))
```

## Session info

```{r}
##########################################################################################
sessionInfo()
```
