---
title: "skylineOutputProcessing"
author: "Christopher Hughes"
format: html
---

## Details

This document describes processing of skyline data from a report export.

## Setting up the environment

These are packages you will need for this notebook. For exact versions used, please refer to the session info at the bottom of this notebook.

```{r}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
library()
```

I want to set a base directory that we can use as a working space.

```{r}
##########################################################################################
baseRepository = 'E:/requests/1243/260211_1243_qtrap'
```

## Data processing

Read in data from Skyline and process to a more readable output.

```{r}
##########################################################################################
##read in the skyline output
##set the molecules of interest
moleculesOfInterest = c('ot82','tyrosine_ilis')
#
skyData = read_csv(paste(baseRepository,'/260211_1243_sampleAnalysis_skylineReport.csv',sep=''),
                   show_col_types = FALSE) %>%
  dplyr::filter(Molecule %in% moleculesOfInterest) %>%
  dplyr::select(Molecule, `Product Mz`, `Replicate Name`, Area) %>%
  dplyr::rename(molecule = Molecule,
                fragment = `Product Mz`,
                sampleName = `Replicate Name`,
                area = Area) %>%
  dplyr::mutate(moleculeName = paste(molecule,'_',fragment, sep='')) %>%
  dplyr::select(-molecule, -fragment) %>%
  dplyr::filter(!grepl('blank', sampleName) & !grepl('ssc', sampleName)) %>%
  dplyr::mutate(sampleName_edit = ifelse(grepl('\\(', sampleName), gsub(' \\([0-9]\\)', '', sampleName), paste(sampleName,'_rep1',sep=''))) %>%
  dplyr::mutate(sampleName_edit = ifelse(grepl('rep1',sampleName_edit), sampleName_edit, paste(sampleName_edit,'_rep2',sep=''))) %>%
  dplyr::mutate(sampleName = sampleName_edit) %>%
  dplyr::select(-sampleName_edit) %>%
  tidyr::pivot_wider(id_cols = sampleName, names_from = moleculeName, values_from = area) %>%
  dplyr::mutate(ot82_frag1_norm = (ot82_300.08 / tyrosine_ilis_145.1)*10000,
                ot82_frag2_norm = (ot82_272.08 / tyrosine_ilis_145.1)*10000)

#output the data
write_tsv(skyData, paste(baseRepository, '/260211_1243_sampleAnalysis_skylineReportParsed.tsv',sep=''))

```
