---
title: "Fragpipe Results Processing"
author: "Christopher Hughes"
format: html
editor: visual
---

## Details

This document describes processing of proteomics data from Fragpipe to compile a final protein set.

## Setting up the environment

These are packages you will need for this notebook. For exact versions used, please refer to the session info at the bottom of this notebook.

```{r}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
```

I want to set a base directory that we can use as a working space.

```{r}
##########################################################################################
baseRepository = 'D:/requests/1178/240125_orbitrapLumos'
```

## Data processing

Read in data from DIA-NN and process with the IQ package.

```{r}
##########################################################################################
##compile the list of directory paths
folderNames = c('fragpipeSample1','fragpipeSample2','fragpipeSample3','fragpipeSample4')


##process through the files
for (i in 1:length(folderNames)){
  #
  rawTemp = read_tsv(paste(baseRepository,'/', folderNames[i], '/protein.tsv',sep=''),show_col_types = FALSE) %>%
    dplyr::select("Protein ID":"Length", "Coverage", "Protein Probability", "Total Peptides":"Razor Spectral Count") %>%
    dplyr::rename(accession  = "Protein ID",
                  uniprot = "Entry Name",
                  gene = "Gene",
                  length = "Length",
                  percentCoverage = "Coverage",
                  proteinProbability = "Protein Probability",
                  totalPeptides = "Total Peptides",
                  uniquePeptides = "Unique Peptides",
                  razorPeptides = "Razor Peptides",
                  totalSpectralCount = "Total Spectral Count",
                  uniqueSpectralCount = "Unique Spectral Count",
                  razorSpectralCount = "Razor Spectral Count") %>%
    dplyr::arrange(-totalSpectralCount)
  #
  saveRDS(rawTemp, paste(baseRepository,'/', folderNames[i], '/proteinProcessed.rds',sep=''))
  write.table(rawTemp, paste(baseRepository,'/', folderNames[i], '/proteinProcessed.tsv',sep=''),
              col.names = TRUE, row.names = FALSE, quote = FALSE, sep = '\t')
}

```

Done.

## Session info

```{r}
##########################################################################################
##session info
sessionInfo()
```



