---
title: "diannDataProcessing"
author: "Christopher Hughes"
format: html
editor: visual
---

## Details

This document describes processing of proteomics data from DIANN to compile a final protein set.

## Setting up the environment

These are packages you will need for this notebook. For exact versions used, please refer to the session info at the bottom of this notebook.

```{r}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
library('iq')
library('DEqMS')
library('RColorBrewer')
library('ggrepel')
```

I want to set a base directory that we can use as a working space.

```{r}
##########################################################################################
baseRepository = 'E:/requests/1192/240404_orbitrapLumos'
```

## Data processing

Read in data from DIA-NN and process with the IQ package.

```{r}
##########################################################################################
##first we can filter and parse the report data into a better format for IQ
dia1 = read_tsv(paste(baseRepository, '/diannDia1/report.tsv',sep=''), show_col_types = FALSE) %>%
  dplyr::mutate(File.Name = sub('.*240412_1193_NRC_sample(.*)_dia[123]\\.raw', '\\1', File.Name)) %>%
  dplyr::filter(!grepl(';', Protein.Group))
#
dia2 = read_tsv(paste(baseRepository, '/diannDia2/report.tsv',sep=''), show_col_types = FALSE) %>%
  dplyr::mutate(File.Name = sub('.*240412_1193_NRC_sample(.*)_dia[123]\\.raw', '\\1', File.Name)) %>%
  dplyr::filter(!grepl(';', Protein.Group))
#
dia3 = read_tsv(paste(baseRepository, '/diannDia3/report.tsv',sep=''), show_col_types = FALSE) %>%
  dplyr::mutate(File.Name = sub('.*240412_1193_NRC_sample(.*)_dia[123]\\.raw', '\\1', File.Name)) %>%
  dplyr::filter(!grepl(';', Protein.Group))
#
raw = rbind(dia1,dia2,dia3)

##save the data
write.table(raw, paste(baseRepository,'/dataProcessing/combinedFilteredReportAnnotated.tsv',sep=''), 
            col.names = TRUE, row.names = FALSE, quote = FALSE, sep = '\t')

##run the IQ analysis
iqResult = process_long_format(paste(baseRepository,'/dataProcessing/combinedFilteredReportAnnotated.tsv',sep=''), 
                           output_filename = paste(baseRepository,'/dataProcessing/combinedFilteredReportAnnotatedIq.tsv',sep=''),
                           annotation_col = 'Genes',
                           normalization = 'median',
                           filter_double_less = c("Lib.Q.Value" = "0.01", "Lib.PG.Q.Value" = "0.01"),
                           peptide_extractor = function(x) gsub("[0-9].*$", "", x))

##read in the human index
annotationDb = readRDS(paste(baseRepository, '/databases/2024-04-21-reviewed-UP000005640-spikein.fasta_index_trypticPeps.rds',sep='')) %>%
  dplyr::select(-species) %>%
  dplyr::filter(!grepl('sp\\|V.*',gene))


##parse the IQ result
proData = read_tsv(paste(baseRepository,'/dataProcessing/combinedFilteredReportAnnotatedIq.tsv',sep=''), show_col_types = FALSE) %>%
  dplyr::rename(accession = 'Protein.Group') %>%
  dplyr::select(-Genes) %>%
  dplyr::left_join(annotationDb)


##save the data
saveRDS(proData, paste(baseRepository, '/dataProcessing/parsedProteinData.rds', sep = ''))
write.table(proData, paste(baseRepository, '/dataProcessing/parsedProteinData.tsv', sep = ''), 
            col.names = TRUE, row.names = FALSE, quote = FALSE, sep = '\t')
```

