---
title: "Theoretical peptides"
author: "Christopher Hughes"
format: html
editor: visual
---

## Details

This document describes processing a fasta database to get a count for the number of theoretical peptides.

## Setting up the environment

These are packages you will need for this notebook. For exact versions used, please refer to the session info at the bottom of this notebook.

```{r}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
library('Peptides')
library('OrgMassSpecR')
library('Biostrings')
```

I want to set a base directory that we can use for our analysis.

```{r}
##########################################################################################
baseRepository = 'E:/requests/1198/240603_orbitrapLumos'
```

## Data processing

Process the contaminants database.

```{r}
##########################################################################################
##crap database processing
fastaDb = readAAStringSet(paste(baseRepository,'/databases/crap_2024_01_15.fasta',sep=''))

##define an index for each entry
fastaIndex = tibble('metadata' = names(fastaDb)) %>%
  dplyr::mutate(accession = paste('sp|CONT', seq(1,nrow(.),1), '|', sub('sp\\|(.*)\\|','\\1', metadata), ' GN=', sub('sp\\|(.*)_.*','\\1', metadata), sep=''))
fastaDbEdited = fastaDb
names(fastaDbEdited) = fastaIndex$accession
toKeep = data.frame(name = names(fastaDbEdited)[which(!grepl('HUMAN', names(fastaDbEdited)))])
fastaSubset = fastaDbEdited[toKeep$name]


##make an index that matches the human one
fastaIndexMatch = fastaIndex %>%
  dplyr::filter(!grepl('HUMAN',accession)) %>%
  dplyr::rename(parsed = 'accession') %>%
  dplyr::mutate(accession = sub('sp\\|(.*)\\|.*','\\1',parsed),
                gene = sub('.*GN\\=(.*)','\\1',parsed),
                length = width(fastaSubset)) %>%
  dplyr::mutate(species = 'contaminant') %>%
  dplyr::select(accession, gene, length, species)

##save this index  
saveRDS(fastaIndexMatch, paste(baseRepository, '/databases/crap_2024_01_15_noHuman_index.rds',sep=''))

##loop through the database entries and calculate tryptic peptide numbers
trypticPeptides = vector()
detectableLength = vector()
for (i in 1:length(fastaSubset)){
  #get the protein sequence
  aaSeq = as.character(fastaSubset[[i]])
  #digest the protein sequence...sometimes there are weird amino acids, so we assign them random masses
  seqDigest = OrgMassSpecR::Digest(aaSeq, enzyme = 'trypsin', missed = 0, custom = list(code = c('X','U','Z','B'), mass = c(50, 60, 70, 80)))
  #calculate the total length of the protein
  seqDigest$pepLength = (seqDigest$stop - seqDigest$start) + 1
  #subset the set of digested peptides based on what we normally detect in an MS experiment
  seqDigestSub = subset(seqDigest, (seqDigest$pepLength > 5) & (seqDigest$pepLength < 31))
  #add to the tryptic peptides vector
  trypticPeptides = c(trypticPeptides, nrow(seqDigestSub))
  #add to the detectable length vector
  detectableLength = c(detectableLength, sum(seqDigestSub$pepLength, na.rm = TRUE))
  #output a message
  message(paste('Finished ', i, ' proteins.', sep = ''))
}


##add the output to our previously generated index
fastaIndexMatch$detectablePeptides = trypticPeptides
fastaIndexMatch$detectableLength = detectableLength


##save the file
saveRDS(fastaIndexMatch, paste(baseRepository, '/databases/crap_2024_01_15_noHuman_index_trypticPeps.rds',sep=''))
writeXStringSet(fastaSubset, paste(baseRepository, '/databases/crap_2024_01_15_noHuman.fasta',sep=''))
```

Go to FragPipe and download the human database appended with this new contaminant one. Uncheck the boxes for decoys and contaminants. Add the contaminant database generated above as a spike-in. After that, we will process the downloaded file using the code below.

Process the fasta database.

```{r}
##########################################################################################
##read in the fasta file
fastaFileName = '2024-06-04-reviewed-UP000005640-spikein.fasta'
fastaDb = readAAStringSet(paste(baseRepository,'/databases/', fastaFileName, sep=''))


##the snippet below
fastaIndex = tibble('metadata' = names(fastaDb)) %>%
  dplyr::mutate(length = width(fastaDb)) %>%
  #dplyr::filter(!grepl('contam_sp', metadata) & !grepl('rev_sp', metadata)) %>%
  dplyr::mutate(accession = sub(".*[sptr]\\|(.*)\\|.*$", "\\1", metadata)) %>%
  dplyr::mutate(gene = ifelse(grepl('GN=', metadata), sub(".*GN=(.*) [PE].*$", "\\1", metadata), NA)) %>%
  dplyr::mutate(species = ifelse(grepl('sapiens', metadata), 'human', 
                             ifelse(grepl('\\=9606', metadata), 'human',
                                    ifelse(grepl('musculus', metadata), 'mouse', 'other')))) %>%
  dplyr::select(accession, gene, length, species)

##save this file as a general index
saveRDS(fastaIndex, paste(baseRepository, '/databases/',fastaFileName,'_index.rds',sep=''))



##loop through the database entries and calculate tryptic peptide numbers
trypticPeptides = vector()
detectableLength = vector()
for (i in 1:length(fastaDb)){
  #get the protein sequence
  aaSeq = as.character(fastaDb[[i]])
  #digest the protein sequence...sometimes there are weird amino acids, so we assign them random masses
  seqDigest = OrgMassSpecR::Digest(aaSeq, enzyme = 'trypsin', missed = 0, custom = list(code = c('X','U','Z','B'), mass = c(50, 60, 70, 80)))
  #calculate the total length of the protein
  seqDigest$pepLength = (seqDigest$stop - seqDigest$start) + 1
  #subset the set of digested peptides based on what we normally detect in an MS experiment
  seqDigestSub = subset(seqDigest, (seqDigest$pepLength > 5) & (seqDigest$pepLength < 31))
  #add to the tryptic peptides vector
  trypticPeptides = c(trypticPeptides, nrow(seqDigestSub))
  #add to the detectable length vector
  detectableLength = c(detectableLength, sum(seqDigestSub$pepLength, na.rm = TRUE))
  #output a message
  message(paste('Finished ', i, ' proteins.', sep = ''))
}


##add the output to our previously generated index
fastaIndex$detectablePeptides = trypticPeptides
fastaIndex$detectableLength = detectableLength


##save the file
saveRDS(fastaIndex, paste(baseRepository, '/databases/',fastaFileName,'_index_trypticPeps.rds',sep=''))
```

After the database is parsed, you can add decoys in FragPipe. If you are going to use it with DIA-NN, there is no need to add decoys. Finished here for now.

## Session info

```{r}
##########################################################################################
sessionInfo()
```
