---
title: "fragpipeOutputProcessing"
author: "Christopher Hughes"
format: html
editor: visual
---

## Details

This document describes processing the output from FragPipe.

## Setting up the environment

These are packages you will need for this notebook. For exact versions used, please refer to the session info at the bottom of this notebook.

```{r}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
```

I want to set a base directory that we can use for our analysis.

```{r}
##########################################################################################
baseRepository = 'E:/requests/1178/240911_1178_qexactive'
```

## Data processing

Process the FragPipe data for all of the samples.

```{r}
##########################################################################################
##read in the fasta index
annotationDb = readRDS(paste(baseRepository, '/databases/2024-09-13-reviewed-UP000009136-spikein.fasta_index_trypticPeps.rds',sep=''))

##read in the peptide result file
sampleList = c('Neg1','Neg2','Neg3','Neg4','Neg5','Palm1','Palm2','Palm3','Palm4','Palm5')
foldersToProcess = c(paste('',sampleList,sep=''))
peptideOutput = tibble()
for (i in 1:length(foldersToProcess)){
  proTemp = read_tsv(paste(baseRepository,'/fragpipeNoMod/',foldersToProcess[i],'/peptide.tsv',sep=''), show_col_types = FALSE) %>%
    dplyr::filter(grepl('DHE3',Gene)) %>%
    dplyr::filter(grepl('C',Peptide)) %>%
    dplyr::select(Peptide, Intensity, `Assigned Modifications`) %>%
    dplyr::rename(peptide = 'Peptide',
                  intensity = 'Intensity',
                  modifications = 'Assigned Modifications') %>%
    dplyr::mutate(sample = sampleList[i])
  #
  peptideOutput = rbind(peptideOutput, proTemp)
}

##shift the data
peptideWide = peptideOutput %>%
  tidyr::pivot_wider(id_cols = c('peptide','modifications'), names_from = 'sample', values_from = 'intensity')

##save the file
write.table(peptideWide, paste(baseRepository,'/fragpipeNoMod/peptideProcessed.tsv',sep=''),
            quote = FALSE, sep = '\t', col.names = TRUE, row.names = FALSE)
```

## Session info

```{r}
##########################################################################################
sessionInfo()
```
