---
title: "Drug compound optimization"
author: "Christopher Hughes"
format: html
---

## Details

This document describes processing the output from Skyline related to optimization of drug compounds.

## Setting up the environment

These are packages you will need for this notebook. For exact versions used, please refer to the session info at the bottom of this notebook.

```{r}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
library('ggpmisc')
library('pastecs')
```

I want to set a base directory that we can use for our analysis.

```{r}
##########################################################################################
baseRepository = 'E:/requests/1205/241107_1205_qexactive'
```

## Data processing

The raw data have already been processed and integrated in Skyline. From here, we will use these data for downstream analyses.

```{r}
##########################################################################################
##read in the data from Skyline
standards = rep(c(500,250,125,62.5,31.25,15.625),each = 3) #in uM
skyRaw = read_csv(paste(baseRepository,'/skyline/setA_moleculeTransitionResults.csv',sep=''),show_col_types = FALSE) %>%
  dplyr::filter(`Product Mz` == 111.01 | `Product Mz` == 113.02) %>%
  dplyr::filter(!grepl('qc_',`Replicate Name`) & !grepl('unknown',`Replicate Name`)) %>%
  dplyr::select(`Replicate Name`, `Product Mz`, Area, Background, `Analyte Concentration`) %>%
  dplyr::rename(sampleName = `Replicate Name`,
                fragmentMass = `Product Mz`,
                signal = 'Area',
                noise = 'Background',
                concentration = `Analyte Concentration`) %>%
  dplyr::mutate(sampleType = ifelse(fragmentMass == 111.01, 'sample', 
                                    ifelse(grepl('blank',sampleName),'blank','ilis')),
                sampleSet = ifelse(grepl('setA',sampleName),'setA',
                                   ifelse(grepl('setB',sampleName),'setB',
                                          ifelse(grepl('setC',sampleName),'setC','blank')))) %>%
  dplyr::mutate(sampleReplicate = sub('.*_rep(.*)','\\1',sampleName)) %>%
  dplyr::mutate(sampleNumber = sub('set[ABC]_(.*)_.*','\\1',sampleName)) %>%
  dplyr::filter(!grepl('blank',sampleName))


##reshape the data
skyWide = skyRaw %>%
  tidyr::pivot_wider(id_cols = c('sampleName','sampleSet','sampleNumber','sampleReplicate'), names_from = 'sampleType', values_from = 'signal') %>%
  dplyr::mutate(relativeToIlis = sample/ilis)


#save the data
saveRDS(skyWide, paste(baseRepository,'/dataProcessing/dataset_processedStandardData.rds',sep=''))
write.table(skyWide, paste(baseRepository,'/dataProcessing/dataset_processedStandardData.tsv',sep=''),
            sep = '\t', quote = FALSE, row.names = FALSE)
```

Calculate the matrix effect (ME), recovery (RE), and process efficiency (PE).

```{r}
##########################################################################################
##read in the processed data
standards = rep(c(500,250,125,62.5,31.25,15.625),each = 3) #in uM
meData = readRDS(paste(baseRepository,'/dataProcessing/dataset_processedStandardData.rds',sep='')) %>%
  dplyr::select(sampleNumber, sampleSet, sampleReplicate, relativeToIlis) %>%
  tidyr::pivot_wider(id_cols = c('sampleNumber','sampleReplicate'), names_from = 'sampleSet', values_from = 'relativeToIlis') %>%
  dplyr::mutate(ME = (setC / setA)*100,
                RE = (setB / setC)*100,
                PE = (ME * RE)/100) %>%
  dplyr::mutate(standardConcentration = standards) %>%
  dplyr::group_by(sampleNumber, standardConcentration) %>%
  dplyr::summarise(setA = mean(setA), setB = mean(setB), setC = mean(setC), ME = mean(ME), RE = mean(RE), PE = mean(PE))
meData

#
write.table(meData, paste(baseRepository,'/dataProcessing/dataset_efficiencyData.tsv',sep=''),
            sep = '\t', quote = FALSE, row.names = FALSE)
```

Plot the different standard curves.

```{r}
##########################################################################################
##read in the processed data
standards = rep(c(500,250,125,62.5,31.25,15.625),each = 3) #in uM
stdData = readRDS(paste(baseRepository,'/dataProcessing/dataset_processedStandardData.rds',sep='')) %>%
  dplyr::select(sampleNumber, sampleSet, sampleReplicate, relativeToIlis) %>%
  dplyr::mutate(sampleConcentration = rep(standards,3))


##setA
dat = dplyr::filter(stdData, grepl('setA',sampleSet))
ggplot(dat, aes(sampleConcentration, relativeToIlis)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2")),rr.digits = 6) +
  geom_point(shape = 21, stroke = 1, size = 2) + 
  labs(x = 'Standard Concentration (uM)', y = 'Area Relative to Internal Standard', title = 'Set A Linearity') +
  theme_classic()
ggsave(paste(baseRepository,'/dataProcessing/point_standardCurve_setA.png',sep=''),
       width = 6, height = 6)

##setB
dat = dplyr::filter(stdData, grepl('setB',sampleSet))
ggplot(dat, aes(sampleConcentration, relativeToIlis)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2")),rr.digits = 6) +
  geom_point(shape = 21, stroke = 1, size = 2) + 
  labs(x = 'Standard Concentration (uM)', y = 'Area Relative to Internal Standard', title = 'Set B Linearity') +
  theme_classic()
ggsave(paste(baseRepository,'/dataProcessing/point_standardCurve_setB.png',sep=''),
       width = 6, height = 6)

##setC
dat = dplyr::filter(stdData, grepl('setC',sampleSet))
ggplot(dat, aes(sampleConcentration, relativeToIlis)) +
  stat_poly_line() +
  stat_poly_eq(use_label(c("eq", "R2")),rr.digits = 6) +
  geom_point(shape = 21, stroke = 1, size = 2) + 
  labs(x = 'Standard Concentration (uM)', y = 'Area Relative to Internal Standard', title = 'Set C Linearity') +
  theme_classic()
ggsave(paste(baseRepository,'/dataProcessing/point_standardCurve_setC.png',sep=''),
       width = 6, height = 6)
```

Use these equations to determine the concentration of the unknowns.

```{r}
##########################################################################################
##get the unknown data
dilutionFactor = 200
unkData = read_csv(paste(baseRepository,'/skyline/setA_moleculeTransitionResults.csv',sep=''),show_col_types = FALSE) %>%
  dplyr::filter(`Product Mz` == 111.01 | `Product Mz` == 113.02) %>%
  dplyr::filter(grepl('unknown',`Replicate Name`)) %>%
  dplyr::select(`Replicate Name`, `Product Mz`, Area, Background, `Analyte Concentration`) %>%
  dplyr::rename(sampleName = `Replicate Name`,
                fragmentMass = `Product Mz`,
                signal = 'Area',
                noise = 'Background',
                concentration = `Analyte Concentration`) %>%
  dplyr::mutate(sampleType = ifelse(fragmentMass == 111.01, 'sample', 
                                    ifelse(grepl('blank',sampleName),'blank','ilis')),
                sampleSet = ifelse(grepl('n1',sampleName),'unknown1',
                                   ifelse(grepl('n2',sampleName),'unknown2',
                                          ifelse(grepl('n3',sampleName),'unknown3','blank')))) %>%
  dplyr::mutate(sampleReplicate = sub('.*_rep(.*)','\\1',sampleName)) %>%
  dplyr::mutate(sampleNumber = sub('set[ABC]_(.*)_.*','\\1',sampleName)) %>%
  tidyr::pivot_wider(id_cols = c('sampleName','sampleSet','sampleNumber','sampleReplicate'), names_from = 'sampleType', values_from = 'signal') %>%
  dplyr::mutate(relativeToIlis = sample/ilis) %>%
  dplyr::mutate(concentration = (((relativeToIlis - 0.0836)/0.024)*dilutionFactor)/1000) #this is in nM, so need to adjust by 1000

#
unkSummary = unkData %>%
  dplyr::group_by(sampleSet) %>%
  dplyr::summarise(concentrationMean = mean(concentration), concentrationSD = sd(concentration))
#
unkData
unkSummary
```

Calculate the original concentrations for the standards based on the standard curve data.

```{r}
##########################################################################################
##get the standard data
standards = rep(c(500,250,125,62.5,31.25,15.625),each = 3) #in uM
stdData = readRDS(paste(baseRepository,'/dataProcessing/dataset_processedStandardData.rds',sep='')) %>%
  dplyr::select(sampleNumber, sampleSet, sampleReplicate, relativeToIlis) %>%
  dplyr::mutate(sampleConcentration = rep(standards,3)) %>%
  dplyr::filter(sampleSet == 'setB') %>%
  dplyr::mutate(calcConcentration = (relativeToIlis - 0.0836)/0.024) %>%
  dplyr::mutate(acc = (calcConcentration/sampleConcentration)*100) %>%
  dplyr::group_by(sampleNumber, sampleSet, sampleConcentration) %>%
  dplyr::summarise(mean_acc = mean(acc, na.rm = TRUE), sd_acc = sd(acc, na.rm = TRUE)) %>%
  dplyr::mutate(cv = (sd_acc/mean_acc)*100)

#
write.table(stdData, paste(baseRepository,'/dataProcessing/dataset_accuracyData.tsv',sep=''),
            sep = '\t', quote = FALSE, row.names = FALSE)
```























