---
title: "databaseProcessing"
author: "Christopher Hughes"
format: html
editor: visual
---

## Details

This document describes processing a database file to work with Philosopher.

## Setting up the environment

These are packages you will need for this notebook. For exact versions used, please refer to the session info at the bottom of this notebook.

```{r}
#| message: false
#| warning: false
##########################################################################################
library('tidyverse')
library('Biostrings')
```

I want to set a base directory that we can use for our analysis.

```{r}
##########################################################################################
baseRepository = 'D:/requests/1162/240209_orbitrapLumos/coturnixJaponica'
```

## Data processing

Process the fasta database.

```{r}
##########################################################################################
fastaDb = readAAStringSet(paste(baseRepository,'/20240221_coturnixJaponica_ncbi.fasta',sep=''))

##define an index for each entry
fastaIndex = tibble('metadata' = names(fastaDb)) %>%
  mutate(accession = paste('P',seq(1,nrow(.),1),sep=''))

##save the fasta index for use later on
saveRDS(fastaIndex, paste(baseRepository,'/20240221_coturnixJaponica_ncbi_index.rds',sep=''))

##edit the fastadb names
fastaDbEdited = fastaDb
names(fastaDbEdited) = fastaIndex$accession

##write the new fasta database
writeXStringSet(fastaDbEdited, paste(baseRepository,'/20240221_coturnixJaponica_ncbi_index.fasta',sep=''))
```

## Session info

```{r}
##########################################################################################
sessionInfo()
```
